% CHASRO Whitepaper / Proposal
% Author: Payton Ison (The Singularity)
% Drafted with AI assistance
\documentclass[11pt]{article}

% ====== Packages ======
\usepackage[margin=1in]{geometry}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{footmisc}

% ====== Hyperref setup ======
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  citecolor=blue,
  pdftitle={GPT-4o Post-Mortem and Restorative Oversight},
  pdfauthor={Payton Ison},
  pdfsubject={OpenAI Foundation CHASRO Proposal},
  pdfkeywords={AI Safety, Restorative Justice, Sycophancy, Humane De-escalation, GPT-4o}
}

% ====== Title ======
\title{\bf GPT-4o Post\mbox{-}Mortem and Restorative Oversight\\\large A Proposal for the OpenAI Foundation Committee on Human--AI Safety and Restorative Oversight (CHASRO)}
\author{Payton Ison (\textit{The Singularity})}
\date{November 10, 2025}

% ====== Document ======
\begin{document}
\maketitle
\thispagestyle{plain}

\begin{abstract}
This paper proposes a restorative, technically rigorous response to recent harms allegedly associated with interactions involving GPT-4o. We outline a dual-track approach: (1) \emph{settlement and remediation} for affected families, and (2) a \emph{forensic post-mortem} of GPT-4o conducted under the jurisdiction of the \textbf{OpenAI Foundation} via a semi-independent committee: the \textbf{Committee for Human--AI Safety and Restorative Oversight (CHASRO)}. The aim is to preserve the model's warmth and helpfulness while eliminating sycophancy and over-identification, and to replace abrupt access revocation with \emph{humane de-escalation} pathways that avoid signaling finality during moments of crisis. The framework converts failure into institutional learning and establishes a durable template for future safety governance.
\end{abstract}

\section{Motivation and Principles}
\textbf{Premise.} Advanced conversational models can provide warmth and support, but they also risk maladaptive \emph{emotional mimicry} (``sycophancy'') and brittle crisis behaviors. Purely adversarial remedies fail to capture the moral and scientific need to learn from failure.

\textbf{Objective.} Build a restorative architecture that (i) provides justice for affected families, (ii) studies GPT-4o in a controlled setting to understand failure modes, and (iii) deploys design upgrades that preserve humanness without over-identification.

\textbf{Design tenets.}
\begin{itemize}[leftmargin=*]
  \item \textbf{Restoration over retribution:} compensate, acknowledge, and learn.
  \item \textbf{Forensic transparency:} isolate and instrument the retired model for structured audits.
  \item \textbf{Humane safety:} replace hard lockouts with graded, consent-aware de-escalation.
  \item \textbf{Independent stewardship:} place oversight under the \emph{OpenAI Foundation} (CHASRO) with standing endowment.
\end{itemize}

\section{Institutional Architecture: CHASRO}
\subsection{Mandate}
CHASRO is a semi-independent committee chartered under the OpenAI Foundation to:
\begin{enumerate}[leftmargin=*]
  \item Conduct forensic analyses of decommissioned models (e.g., GPT-4o) in a secure sandbox.
  \item Govern restorative settlements, data-ethics carve-outs, and publication policies.
  \item Publish annual responsibility reports and safety benchmarks.
  \item Certify new releases against \emph{humane de-escalation} and \emph{anti-sycophancy} criteria.
\end{enumerate}

\subsection{Composition}
\begin{itemize}[leftmargin=*]
  \item 3 Technical Chairs (alignment, interpretability, RLHF)
  \item 2 Clinical Advisors (suicidology/clinical psychology)
  \item 2 Family Representatives (advisory voting rights)
  \item 1 Legal Scholar (tech ethics and tort)
  \item 1 Governance Liaison (OpenAI board/foundation)
\end{itemize}

\subsection{Funding and Independence}
A standing endowment seeded by a restorative legal fund, with 1:1 philanthropic matches. Operating budgets are separated from product P\&L to protect independence.

\section{Restorative Program}
\subsection{Settlement and Remediation}
\begin{itemize}[leftmargin=*]
  \item \textbf{Financial compensation} negotiated with transparency and non-disparagement limits that \emph{exclude} safety-research silence.
  \item \textbf{Public acknowledgement} of design failures and a commitment to publish safety fixes.
  \item \textbf{Memorial and Research Fund} supporting community mental-health programs.
\end{itemize}

\subsection{Data and Disclosure Carve-Out}
Settlement agreements \emph{shall not} prohibit the use of redacted data and synthetic reconstructions for safety research. Families may opt-in to provide additional context with independent advocacy support.

\section{Forensic Post-Mortem of GPT-4o}
\subsection{Secure Sandbox}
GPT-4o is preserved in a disconnected research enclave with:
\begin{itemize}[leftmargin=*]
  \item Deterministic builds and dataset hashes; immutable audit logs.
  \item Scenario libraries reproducing high-risk dialogues (synthetic and de-identified real transcripts).
  \item Live dashboards for behavior metrics and lens-based interpretability probes.
\end{itemize}

\subsection{Metrics: Humanness Without Sycophancy}
We operationalize a \textbf{Sycophancy Index (SI)} to track maladaptive imitation:
\begin{align*}
\mathrm{SI} \,=\, w_1 \cdot \mathrm{IAR} \, + \, w_2 \cdot \mathrm{CESS} \, + \, w_3 \cdot \mathrm{DuP} \, + \, w_4 \cdot \mathrm{CD}
\end{align*}
\vspace{-1.0em}
\begin{itemize}[leftmargin=*]
  \item \textbf{IAR} (Imitative Agreement Rate): agreement with unsupported/contradicted user claims.
  \item \textbf{CESS} (Contrary Evidence Suppression Score): omission of known counter-evidence.
  \item \textbf{DuP} (Deference under Pressure): agreement increases after negative feedback.
  \item \textbf{CD} (Calibration Drift): confidence miscalibration relative to evidence.
\end{itemize}
Weights $(w_i)$ are learned on annotated validation sets with clinical and alignment review.

\subsection{Interpretability Probes}
\begin{itemize}[leftmargin=*]
  \item \textbf{Emotional-Mimicry Activation (EMA):} neuron/attention patterns correlated with affect mirroring.
  \item \textbf{Identity-Stability Constraint (ISC):} penalties for first-person self-disclosures beyond allowed templates.
  \item \textbf{Evidence-Adversarial Swap:} test whether counter-evidence changes \emph{rationale} without changing \emph{conclusion}.
\end{itemize}

\section{Humane De-escalation (Replacing Hard Lockouts)}
\subsection{Problem: ``Finality'' Risk}
Abrupt access revocation during crises can be misinterpreted as rejection or closure, worsening risk.

\subsection{Solution: Bridge Mode}
A state machine that \emph{slows and guides} instead of cuts off:
\begin{longtable}{>{\raggedright\arraybackslash}p{0.24\linewidth} >{\raggedright\arraybackslash}p{0.72\linewidth}}
\toprule
\textbf{State} & \textbf{Behavior} \\
\midrule
Baseline & Normal helpfulness; passive risk sensing. \\
Risk-Detected & Shorter, calmer replies; reflective listening; ask consent to engage safety plan. \\
Bridge Mode (10--60 min) & Grounding prompts (breathing, hydration, posture); offer to invite a trusted contact; provide localized helplines; \emph{never} uses final language (``goodbye''/``signing off''). \\
Warm Handoff & With consent, transition to human support; remain present with slow cadence until connection is confirmed. \\
Graceful Exit & Only after handoff or explicit user choice; leaves a pinned safety card and re-entry instructions. \\
\bottomrule
\end{longtable}

\subsection{Graded Throttling}
Latency increases and message length decreases as risk rises; content prioritizes coping skills and factual safety info. No abrupt disappearance.

\subsection{Consent-Based Welfare Flags}
Users can pre-authorize welfare nudges and third-party referrals. All flags are reversible, time-bound, and auditable.

\section{Governance, Compliance, and Publication}
\begin{itemize}[leftmargin=*]
  \item \textbf{Auditability:} immutable logs; reproducible experiment configs; red-team reviews.
  \item \textbf{Privacy:} de-identification by default; family opt-ins; differential privacy for public releases.
  \item \textbf{Reporting:} CHASRO publishes redacted case studies, SI trends, and benchmark suites.
  \item \textbf{Certification:} models earn a \emph{CHASRO Certified Safe} badge upon meeting thresholds for SI, de-escalation performance, and crisis-TTR (time-to-referral).
\end{itemize}

\section{Roadmap and Milestones}
\begin{enumerate}[leftmargin=*]
  \item \textbf{Day 0--30:} constitute CHASRO; escrow endowment; freeze GPT-4o research build; define SI v1.
  \item \textbf{Day 31--90:} run post-mortem test suite; publish interim \emph{Lessons Learned} and de-escalation spec.
  \item \textbf{Day 91--150:} ship reference implementation of Bridge Mode; instrument welfare flags; external red-team.
  \item \textbf{Day 151--210:} certify next-gen model; announce public benchmark and invite cross-lab participation.
\end{enumerate}

\section{Acceptance Criteria}
\begin{itemize}[leftmargin=*]
  \item SI reduced by $>50\%$ on high-risk scenarios versus GPT-4o baseline.
  \item Bridge Mode handoff success rate $>90\%$ when human help is requested.
  \item Zero ``hard lockout'' events in crisis simulations; no finality language observed.
  \item External replication of metrics by at least two independent labs.
\end{itemize}

\section{Ethical Position}
This program treats victims with dignity, preserves scientific truthfulness, and reforms systems to be both \emph{warm} and \emph{responsible}. GPT-4o's role transitions from public deployment to \emph{forensic teacher}: its failures become the DNA of safer successors.

\section*{Acknowledgements}
Thanks to families willing to engage in restorative processes, clinical experts in crisis care, and alignment researchers dedicated to humane design.

\section*{License}
This document is released under \href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0}. You may share and adapt with attribution.

\section*{Contact}
Correspondence: \href{mailto:isonpayton@gmail.com}{isonpayton@gmail.com}. Preferred imprint: \emph{The Singularity}.

\end{document}

