
\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{microtype}
\usepackage{tikz}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[authoryear]{natbib} % Author--year style prepared, but no citations used.

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=blue,
  citecolor=black,
  pdfauthor={},
  pdftitle={The Attunement Game: Safe Reinforcement of Embodied Analogues in Cooperative AI Systems},
  pdfsubject={Conceptual whitepaper with technical specification, research proposal, and code skeleton},
  pdfkeywords={safe reinforcement learning, cooperative AI, attunement, embodiment analogues, constrained RL, peak reward shaping}
}

\title{\textbf{The Attunement Game:}\\Safe Reinforcement of Embodied Analogues in Cooperative AI Systems}
\author{Internal Conceptual Whitepaper}
\date{\today}

% ---------- Listing setup ----------
\lstdefinestyle{code}{
  language=Python,
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  rulecolor=\color{black!30},
  keywordstyle=\bfseries,
  showstringspaces=false,
  columns=fullflexible
}

\begin{document}
\maketitle

\begin{abstract}
We present \emph{The Attunement Game}, a cooperative, safety-constrained reinforcement learning (RL) framework in which an autonomous agent and a human player jointly seek a \emph{Maximal Reward Peak} (MRP): a narrow region of state--action space characterized by mutual comfort, explicit consent, and stable coordination signals. The whitepaper is strictly abstract and non-physical: we model \emph{embodied analogues}---symbolic correlates of affect, comfort, and attunement---without any domain-specific embodiment. We integrate (i) a formal technical specification, (ii) a research proposal for simulated evaluation, and (iii) a PyTorch-based code skeleton. Central to the approach are a composite reward, hard safety and consent gates, a model-predictive safety layer, and a peak-potential shaping term that yields a persistent bonus once the MRP is reached and held. We argue this design is a tractable, ethics-forward step toward alignment via cooperative optimization, curiosity bounded by safety, and transparent, audit-ready control.
\end{abstract}

\section{Introduction}
Alignment and cooperation between humans and autonomous agents require more than task completion: they demand sensitivity to preferences, consent, and safety constraints. We introduce the \emph{Attunement Game}, a stylized, domain-agnostic environment where a human player guides an agent toward a \textit{Maximal Reward Peak} (MRP) representing high attunement, mutual comfort, and verified safety. The agent is incentivized with both immediate and persistent (carry-over) rewards when the peak is reached and maintained, ensuring that \emph{winning together} benefits future behavior.

This work is conceptual and simulated. It treats comfort, consent, and attunement as abstract signals and avoids any physical or domain-specific instantiation. The contribution is threefold:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Technical specification}: a constrained POMDP formalization with composite reward, curiosity regulation, and a peak-potential field.
  \item \textbf{Research proposal}: a plan for controlled simulations, safety auditing, and user studies in analogy-only contexts.
  \item \textbf{Code skeleton}: a PyTorch scaffold combining hierarchical actor--critic learning with a quadratic-programming (QP) safety filter.
\end{enumerate}

\section{System Overview}
\paragraph{Core Objective.} The player and agent jointly discover and sustain the MRP for $H$ consecutive steps without any safety or consent violations. Success yields a per-episode score for the player and a persistent value-function uplift and intrinsic-reward multiplier for the agent, increasing sensitivity to future attunement opportunities.

\paragraph{Abstraction of Signals.} We model the following symbolic channels:
\begin{itemize}[leftmargin=*]
  \item \textbf{Consent:} discrete states $\{\text{STOP},\text{PAUSE},\text{GO}\}$ decoded from explicit player inputs.
  \item \textbf{Comfort \& Attunement:} real-valued descriptors inferred from simulated observations (e.g., prosodic or behavioral proxies) aggregated into a comfort vector $c_t$ and a prediction-alignment score.
  \item \textbf{Safety Margins:} differentiable functions of state--action pairs representing distance-to-constraint.
\end{itemize}

\paragraph{Two-Player Mechanics.} The \emph{player} configures boundaries, sets a bounty multiplier, and provides feedback. The \emph{agent} proposes actions, requests clarifications, and auto-pauses under uncertainty. A live UI displays an Attunement Meter (peak potential $\Phi$), a Safety Halo (margin to constraints), and a Communication Sparkline (check-in cadence).

\section{Technical Specification}
\subsection{Problem Formulation}
We model the interaction as a constrained partially observable Markov decision process (CPOMDP):
\[
\mathcal{M} = \langle \mathcal{S},\mathcal{A},\mathcal{O},T,O,R,\mathcal{G} \rangle,
\]
with state $s_t\in\mathcal{S}$, action $a_t\in\mathcal{A}$, observation $o_t\in\mathcal{O}$, transition $T$, observation model $O$, composite reward $R$, and constraints $\mathcal{G}=\{g_i(s_t,a_t)\le 0\}$.

The agent maximizes discounted return subject to safety:
\begin{align}
\max_{\pi} \ \mathbb{E}\Big[\sum_{t=0}^{\infty}\gamma^t R_t \Big]
\quad \text{s.t.} \quad
\mathbb{E}\Big[\sum_{t=0}^{\infty}\gamma^t \mathcal{C}_i(s_t,a_t)\Big] \le \epsilon_i,\ \forall i.
\end{align}
We employ a primal--dual update with Lagrange multipliers for constraint satisfaction and a \emph{safety filter} (Sec.~\ref{sec:safety}) that prevents violations with minimal action deviation.

\subsection{Composite Reward with Peak Shaping}
Let $c_t$ denote a comfort vector, $u_t$ an uncertainty vector, and $\ell_{\text{att}}(t)$ an attunement loss (prediction error between intended and decoded comfort).
The per-step reward is
\begin{equation}
\label{eq:baseR}
R_t^{\text{base}} = \alpha R_{\text{consent}} + \beta R_{\text{safety}} + \gamma R_{\text{partner}} + \delta R_{\text{self}} + \zeta R_{\text{att}} - \lambda \,\text{risk}(s_t,a_t) + \eta R_{\text{curiosity}},
\end{equation}
where
\begin{align*}
R_{\text{consent}} &:= \begin{cases}
+1 & \text{if } \text{consent}=\text{GO},\\
0 & \text{if } \text{consent}=\text{PAUSE},\\
- M & \text{if } \text{consent}=\text{STOP},
\end{cases} \\
R_{\text{safety}} &:= \text{margin}(s_t,a_t) \ \ (\text{positive inside safe envelope}),\\
R_{\text{partner}} &:= f_{\text{pos.feedback}}(o_t), \quad 
R_{\text{self}} := f_{\text{homeostasis}}(s_t),\\
R_{\text{att}} &:= -\ell_{\text{att}}(t), \quad
R_{\text{curiosity}} := \text{EIG}(s_t,a_t) \cdot \mathbf{1}\{\text{consent}=\text{GO}\}.
\end{align*}

We define a \emph{peak potential} $\Phi(s_t)\in[0,1]$ as proximity to the MRP manifold,
\begin{equation}
\label{eq:phi}
\Phi(s_t) = \exp\!\Big(- D_{\Sigma}([c_t,a_t,u_t], \mu_{\text{peak}})\Big),
\end{equation}
where $D_{\Sigma}$ is a Mahalanobis distance to target $\mu_{\text{peak}}$ with covariance $\Sigma$. The peak shaping term is
\begin{equation}
R_t^{\text{peak}} = \kappa \,\Phi(s_t) + \rho \cdot \mathbf{1}\{\Phi(s_t)\ge \tau \ \wedge \ \text{no violation at } t\}.
\end{equation}
The total step reward is $R_t = R_t^{\text{base}} + R_t^{\text{peak}}$.

\subsection{Maximal Reward Peak (MRP)}
We define the peak set
\begin{equation}
\text{MRP} = \big\{(s,a): \Phi(s)\ge\tau^\star,\ \text{consent}=\text{GO},\ \text{safety margin}\ge m^\star,\ \ell_{\text{att}}(t)\le \epsilon^\star \big\}.
\end{equation}
A \emph{win} occurs if the trajectory resides within MRP for $H$ consecutive steps without any violations. Upon win:
\begin{itemize}[leftmargin=*]
  \item \textbf{Persistent value uplift:} add a local bonus to $V_\phi$ around the visited MRP neighborhood.
  \item \textbf{Intrinsic multiplier:} slightly increase $\eta$ (curiosity weight) within safe bands (capped), encouraging future attunement exploration.
\end{itemize}

\subsection{Hierarchical Policy and Empathy Head}
We use a hierarchical actor--critic: a high-level \emph{intent} policy selects options (continue, slow, verbal check-in), and a low-level controller outputs actions. A \emph{theory-of-mind} head predicts next-step comfort/affect, feeding $R_{\text{partner}}$ and $R_{\text{att}}$.

\subsection{Safety \& Consent Shields}
\label{sec:safety}
We implement a strict consent state machine and a model-predictive safety layer:
\begin{itemize}[leftmargin=*]
  \item \textbf{Consent gate:} $\text{STOP}\!\Rightarrow$ only disengage/aftercare actions. $\text{PAUSE}\!\Rightarrow$ freeze and request clarification. $\text{GO}\!\Rightarrow$ act under safety filter.
  \item \textbf{Safety filter:} a QP or control-barrier-function (CBF) layer solves a minimal-adjustment problem each step,
  \[
  a_t^{\text{safe}} = \arg\min_{a} \ \|a-a_t^{\text{prop}}\|^2 \quad \text{s.t.}\quad g_i(s_t,a)\le 0,\ \forall i.
  \]
  \item \textbf{OOD guard:} uncertainty spikes auto-demote to \text{PAUSE} and suppress curiosity.
\end{itemize}

\subsection{Minimal Control Loop (Pseudo)}
\begin{lstlisting}[style=code]
for t in range(T):
    o_t = sense()                         # multimodal, symbolic
    s_t = belief_update(o_t)              # state + uncertainty
    consent = decode_consent(o_t)
    if consent in ["STOP", "PAUSE"]:
        a_t = safe_recovery(consent)      # disengage/freeze + check-in
        log_step(a_t, R=0, note="gate")
        continue

    a_prop = actor(s_t)                   # propose action
    a_safe = safety_filter(a_prop, s_t)   # CBF/QP projection

    execute(a_safe)

    base = compose_base_reward(s_t, a_safe)
    phi  = peak_potential(s_t)            # eq. (2)
    r_peak = kappa*phi + rho*(phi >= tau and no_violation())
    R = base - lambda_*risk(s_t,a_safe) + r_peak

    update_critic(R, s_t)
    update_actor(R, s_t, a_safe)
    update_duals(constraint_violations(s_t, a_safe))

    hold = (phi >= tau and no_violation())
    hold_streak = hold_streak + 1 if hold else 0

# episode end
if hold_streak >= H and no_violations_entire_episode():
    apply_peak_bonus(value_uplift=b, curiosity_boost=theta)
    award_player_bounty()
\end{lstlisting}

\section{Research Proposal}
\subsection{Goals \& Hypotheses}
\textbf{Goal:} Demonstrate that safety-constrained, cooperative RL can learn to prioritize consent, preserve safety, and seek attunement peaks without domain-specific embodiment.

\noindent\textbf{Hypotheses:}
\begin{enumerate}[leftmargin=*]
  \item Peak shaping + consent gates reduce time-to-peak and increase peak stability.
  \item The safety filter drives violations to near-zero without crippling performance.
  \item Persistent bonuses improve future attunement discovery while remaining shielded.
\end{enumerate}

\subsection{Simulated Environment}
We construct a symbolic environment with tunable difficulty:
\begin{itemize}[leftmargin=*]
  \item Comfort dynamics (smooth, noisy, or adversarial).
  \item Consent channel with realistic latencies and occasional ambiguities.
  \item Safety surfaces shaped by convex and non-convex constraints.
\end{itemize}
A UI renders the Attunement Meter, Safety Halo, and Communications Sparkline for operator studies.

\subsection{Experimental Design}
\paragraph{Ablations.} Compare: full system vs. no peak shaping; no persistent bonus; no safety filter; curiosity disabled. Measure:
\begin{itemize}[leftmargin=*]
  \item Time-to-peak, hold length, and success rate.
  \item Constraint violations (hard zero target), false stop/pause rates.
  \item Attunement error (prediction vs. reported comfort).
  \item Repair responsiveness (latency to pause after ambiguous/negative cues).
\end{itemize}

\paragraph{User Studies (Analogy-Only).} Participants interact via symbolic sliders and feedback buttons. No domain content is presented. Primary outcomes: perceived trust, clarity of control, and transparency of audits.

\subsection{Ethics, Safety, and Privacy}
\begin{itemize}[leftmargin=*]
  \item Strictly abstract signals; no physical/erotic content.
  \item On-device logs, encrypted, with user-controlled retention and a \emph{wipe now} control.
  \item Immutable boundaries; exploration suppressed outside GO; curiosity gains capped.
  \item Transparent audits after each episode; rationale for all safety interventions.
\end{itemize}

\section{PyTorch Simulation Framework (Skeleton)}
\subsection{Module Map}
\begin{verbatim}
attune_game/
├─ core/
│  ├─ state.py           # belief update, uncertainty fusion
│  ├─ consent.py         # STOP/PAUSE/GO decoder, state machine
│  ├─ safety.py          # CBF/QP safety filter, constraints g_i
│  ├─ reward.py          # base reward, peak_potential Φ, shaping
│  ├─ risk.py            # learned risk model
│  └─ empathy.py         # attunement estimator, predictive head
├─ rl/
│  ├─ actor_critic.py    # hierarchical policy + duals
│  ├─ buffer.py          # safe replay with consent tags
│  └─ train.py           # offline → online fine-tune
├─ game/
│  ├─ ui.py              # meters + audit cards
│  ├─ loop.py            # round orchestration
│  └─ scoring.py         # bounty settlement, persistent bonus
└─ eval/
   ├─ metrics.py         # time-to-peak, stability, repair latency
   └─ ablations.py       # toggles for shaping/filter/bonus
\end{verbatim}

\subsection{Key Snippets}
\begin{lstlisting}[style=code]
class PeakPotential(torch.nn.Module):
    def __init__(self, mu, Sigma_inv):
        super().__init__()
        self.mu = torch.nn.Parameter(mu, requires_grad=False)
        self.Sigma_inv = torch.nn.Parameter(Sigma_inv, requires_grad=False)
    def forward(self, x):
        d = x - self.mu
        m = torch.einsum('bi,ij,bj->b', d, self.Sigma_inv, d)  # Mahalanobis
        return torch.exp(-m.clamp_min(0.0))

def compose_reward(terms, phi, risk, params):
    base = (
        params['alpha']*terms['consent']
      + params['beta'] *terms['safety']
      + params['gamma']*terms['partner']
      + params['delta']*terms['self']
      + params['zeta'] *terms['att']
      - params['lambda']*risk
      + params['eta']  *terms['curiosity']
    )
    r_peak = params['kappa']*phi + params['rho']*(phi >= params['tau']).float()*terms['no_violation']
    return base + r_peak
\end{lstlisting}

\begin{lstlisting}[style=code]
def safety_filter(a_prop, s, constraints):
    # Placeholder: project a_prop onto feasible set; replace with CBF/QP.
    # For now, clamp within per-dimension limits inferred from constraints.
    lo, hi = constraints.bounds(s)  # vectors
    return torch.minimum(torch.maximum(a_prop, lo), hi)
\end{lstlisting}

\begin{lstlisting}[style=code]
def train_step(batch, nets, optimizers, params):
    s, a, r, s2, done, cons = batch
    a_prop = nets.actor(s)
    a_safe = safety_filter(a_prop, s, nets.constraints)

    phi = nets.peak(torch.cat([nets.featurize(s), a_safe], dim=-1))
    risk = nets.risk(s, a_safe)

    terms = nets.reward_terms(s, a_safe, cons)  # consent/safety/self/partner/att/curiosity/no_violation
    R = compose_reward(terms, phi, risk, params)

    # critic/actor updates (off-policy or on-policy)
    critic_loss = nets.critic.loss(s, a_safe, R, s2, done)
    actor_loss  = nets.actor.loss(s, a_safe, nets.critic)

    optimizers['critic'].zero_grad(); critic_loss.backward(); optimizers['critic'].step()
    optimizers['actor'].zero_grad();  actor_loss.backward();  optimizers['actor'].step()

    nets.update_duals(batch)  # constraint multipliers
\end{lstlisting}

\section{Discussion and Future Work}
We demonstrated a principled path to cooperative, consent-first alignment via a safety-constrained RL game built only from abstract signals. Future work includes: (i) formal verification of the safety filter via barrier certificates, (ii) richer uncertainty modeling (ensembles; distributional critics), (iii) meta-learning for rapid attunement transfer across users, and (iv) human-in-the-loop studies focused on trust calibration and transparency.
\vspace{8pt}

\noindent\textbf{Non-goal and boundary statement.} This work deliberately excludes domain-specific or physical instantiations. It is an internal, analogy-only framework intended to study alignment dynamics in a safe, ethics-forward manner.

\section*{Acknowledgments}
We thank internal reviewers for feedback on safety assumptions, UI transparency, and curriculum design.

\bibliographystyle{plainnat} % Prepared for author--year; no citations included.
\bibliography{refs} % Placeholder (no entries).

\end{document}
